---
editor_options: 
  markdown: 
    wrap: 72
---

# Data Manipulation

Exploring hierarchy in the data

visualization of order of data

x and y , both are numeric

1.  barplot

2.  spider plot or radar

3.  word cloud

4.  parallel plot

5.  lollipop plot

6.  circular barplot


## Data Manipulation Skills

```{r}
setwd("C:/Users/Hello People/Desktop/AMA SY 21-22 2ND SEM/datanal") # to set working directory
getwd()  # to see which is our working directory
```
### reading data from CSV or excel file
```{r}
#read.csv() function
df= read.csv("iris.csv")
head(df)
summary(df)
str(df)
```

```{r}
#read.xlsx() function
df1= readxl::read_xlsx("iris.xlsx")
head(df1)
summary(df1)
str(df1)
```

```{r}
# if your data is in specific sheet of excel workbook
df2= readxl::read_xlsx("iris.xlsx",sheet="iris")
head(df2)
summary(df2)
str(df2)
```
## Selecting Columns and Rows
```{r}
# for the function below we need tidyverse package
library(tidyverse)
# Selecting column

df_f4= select(df,-5)
head(df_f4)
```

```{r}
# filtering rows
df_setosa= filter(df,Species=="setosa")
head(df_setosa)
dim(df_setosa)

df_virginica= filter(df,Species=="virginica")
head(df_virginica)
dim(df_virginica)
```
## Combining Columns and Rows
```{r}
# gathering columns to make long table
df_long= gather(df,Mes_type,length,1:4)
head(df_long)
```

```{r}
#grouping and summarizing data
# group by
df_sumzd=group_by(df_long,Species)
head(df_sumzd)
```

```{r}
#Summarize
df_sumzd=group_by(df_long,Species,Mes_type) %>% summarise(mean=mean(length),sd=sd(length))
head(df_sumzd)
```
## Basic Bar Plots
```{r}

#p=ggplot(df_sumzd,aes(x=Species,y=mean))+geom_bar()

#p
```

```{r}
# Plot basic barplot- effect of stat= "identity"
p=ggplot(df_sumzd,aes(x=Species,y=mean))+
   geom_bar(stat="identity")
p
```

```{r}
# fill color
p=ggplot(df_sumzd,aes(x=Species,y=mean))+
   geom_bar(stat="identity",fill="red")
p
```

```{r}
p=ggplot(df_sumzd,aes(x=Species,y=mean, fill=Species))+
   geom_bar(stat="identity")
p
```
# Fill Color - mapping to a variable
```{r}

# fill colors- stacked bar plot
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type))+
   geom_bar(stat="identity")
p
```

```{r}
# fill colors- grouped bar plot 
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type))+
   geom_bar(stat="identity",position="dodge")
p
```

```{r}
#Error bars
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type))+
   geom_bar(stat="identity",position="dodge") +
   geom_errorbar(aes(ymin=mean-sd,ymax=mean+sd))

p
```

```{r}
#Error bars width
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type))+
   geom_bar(stat="identity",position="dodge") +
   geom_errorbar(aes(ymin=mean-sd,ymax=mean+sd),width=0.25)

p
```

```{r}
#Error bars width, size,position
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type))+
   geom_bar(stat="identity",position="dodge") +
   geom_errorbar(aes(ymin=mean-sd,ymax=mean+sd),width=0.25,
                 size=1,position= position_dodge(0.9))
p
```

```{r}
#Error bars width,size, position, alpha
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type))+
   geom_bar(stat="identity",position="dodge") +
   geom_errorbar(aes(ymin=mean-sd,ymax=mean+sd),width=0.25,
                 size=1,position= position_dodge(0.9),alpha=0.3)
p
```

```{r}
#Error bars width,size, position, alpha and labels
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type,label=mean))+
   geom_bar(stat="identity",position="dodge") +
   geom_errorbar(aes(ymin=mean-sd,ymax=mean+sd),width=0.25,
                 size=1,position= position_dodge(0.9), alpha=0.3)
p
```

```{r}
#Error bars width,size, position, alpha and labels
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type,label=mean))+
   geom_bar(stat="identity",position="dodge") +
   geom_errorbar(aes(ymin=mean-sd,ymax=mean+sd),width=0.25,
                 size=1,position= position_dodge(0.9), alpha=0.3)+
   geom_text(position=position_dodge(0.9))

p
```

```{r}
#Error bars width,size, position, alpha and labels, label position corrected
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type,label=mean))+
   geom_bar(stat="identity",position="dodge") +
   geom_errorbar(aes(ymin=mean-sd,ymax=mean+sd),width=0.25,
                 size=1,position= position_dodge(0.9), alpha=0.3)+
   geom_text(position=position_dodge(0.9),vjust=-1)

p
```

```{r}
#Error bars width,size, position, alpha and labels, label position more corrected
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type,label=mean))+
   geom_bar(stat="identity",position="dodge") +
   geom_errorbar(aes(ymin=mean-sd,ymax=mean+sd),width=0.25,
                 size=1,position= position_dodge(0.9), alpha=0.3)+
   geom_text(position=position_dodge(0.9),vjust=-0.5, hjust=1.2)
p
```

```{r}
#theme classic
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type,label=mean))+
   geom_bar(stat="identity",position="dodge") +
   geom_errorbar(aes(ymin=mean-sd,ymax=mean+sd),width=0.25,
                 size=1,position= position_dodge(0.9), alpha=0.3)+
   geom_text(position=position_dodge(0.9),vjust=-0.5, hjust=1)+
   theme_classic()
p
```

```{r}
#theme bw
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type,label=mean))+
   geom_bar(stat="identity",position="dodge") +
   geom_errorbar(aes(ymin=mean-sd,ymax=mean+sd),width=0.25,
                 size=1,position= position_dodge(0.9), alpha=0.3)+
   geom_text(position=position_dodge(0.9),vjust=-0.5, hjust=1.1)+
   theme_bw()

p
```

```{r}
# I don't like this plot.
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type))+
   geom_bar(stat="identity",position="dodge") +
   geom_errorbar(aes(ymin=mean-sd,ymax=mean+sd),width=0.25,
                 size=1,position= position_dodge(0.9), alpha=0.3)+
      theme_bw()
p
```

```{r}
# coord flip, this to when x axis labels are very long and overlap each other
p=ggplot(df_sumzd,aes(x=Species,y=mean,fill=Mes_type))+
   geom_bar(stat="identity",position="dodge") +
   geom_errorbar(aes(ymin=mean-sd,ymax=mean+sd),width=0.25,
                 size=1,position= position_dodge(0.9), alpha=0.3)+
   theme_bw()+
   coord_flip()
p
```

#That's it for bar plot.
